### 2025-04-10 tfp2025 3/review-response.md

Traditionally, a *grammar* is a list of nonterminals and collection of rewrite 
rules from the nonterminals to lists of nonterminals and terminals. A grammar
defines a language, namely the set of all strings (lists of nonterminals)
reachable by applying rewrite rules starting from a designated the starting
nonterminal. Note that grammars have more structure than languages. For example,
a string can be generated by a grammar in multiple different ways. Our type
theoretic definition of language correlates, in that respect, more directly to
traditional grammars. 
However, in this paper we define grammars in a slightly different way. We use an
algebraic/expression based definition ... TODO!

A *parser* is usually defined as a function from strings to booleans in theory
or a program which turns a string into a parse tree in practice.

> * One of the main technical contributions of the paper is the treatment
>  of the recursion and derivatives. To compute the derivative, there is
>  an additional mu constructor, introducing an indirection to
>  distinguish between the original language and the recursive calls to
>  the newly computed derivative. Given that the paper is mostly
>  interested in parsing, one might wonder if the indirection is strictly
>  necessary: couldn't the intended semantics (i.e. recursive call to the
>  derivative function of the _original_ grammar) not be passed as an
>  argument to δ₀? If this isn't the case, it might be worth
>  explaining/motivating why the direct solution does not work - I could
>  imagine it causes problems for the termination checker, for instance.

TODO: I don't understand the suggestion, actually. We use the mu to preserve
the original grammar, so passing the derivative of that to `δ₀` does not
help us. We cannot reconstruct the original grammar from the derivative of it.

TODO: The only way we might hope to avoid `mu` is if we take a more
big-step/recursive descent approach with memoization, then it might be possible.
We could explain this in the discussion of possible future work.

@@

One of the main difficulties is that our nullability correctness theorem is
not a one-to-one equivalence. We intentionally exclude parses which take 
unproductive recursive iterations.

Furthermore, I don't think ambiguous grammars are useful in practice (in CS), so
our effort would be better spent developing better tools to check and prove
that our languages are not ambiguous, avoiding the whole issue of having
multiple parse trees.

> * One important reason for separating out nullability and derivatives is
>  that it becomes possible to define the parser by induction on the
>  input string. Otherwise, it is not so clear that parsing will
>  terminate at all – and often does not. I'd suggest clarifying this
>  point halfway through page 6.

Good point. Although, this is not obvious. I need some time to think about this.
Also, this is strongly related to your second major suggestion. If

@@

The nu notation is quite standard. Brzozowski originally used delta-zero, but I
already use that for a different purpose and I think that would also be
similarly confusing.

> * There is quite some work on (generic programming with) fixpoints of
>  indexed functors or mutually recursive families of datatypes. Citing
>  this as an avenue for further work might help substantiate the
>  argument that these ideas can be extended to handle richer grammars.

Yes, in the discussion on expressiveness we should cite indexed data type
descriptions and how they can be used to model mutually recursive data types.
That work does port over directly to our grammars. However, I did not manage to
finish all the proofs yet, because the indexing makes them quite a bit more
complicated.

> ### Typos
> * book-keeping -> bookkeeping
> * a program which can prove for us -> proves whether or not ...
> * condintional -> conditional
> * 'Most practical programming languages' and 'Many practical programming languages' - repitition.
> * allowed to take the fixpoint of -> of which we are allowed to take the
>  fixpoint (never end a sentence with a proposition).
> * D But (missing punctuation mark)

@@

> In conclusion, putting these points in consideration, my evaluation of
> this paper is in-between 0 and 1. I am slightly positive but not that
> positive to score 1.

We can add that paper to the related work section, it will be similar to the
paper on the left-corner transformation.

Additionally, we should clarify that none of the proofs or their formalization
are novel. Instead, our approach is novel and useful because it is simpler and
more direct than traditional approaches (our code is roughly one tenth of the
size of Firsov and Uustalu's code: ~400 vs. 4159 lines).

@@

> In addition, I would strongly suggest that the author change
> the title, because the present paper is not the first one that shows
> mechanized formalization of context-free grammars.

TODO: With "Type Theoretically" I intended to mean that this is a very specific
perspective on how to formalize, so we don't just present a formalization.

TODO: Consider changing title, e.g. "Towards a Type Theoretic Treatment of Context-Free Languages"

> 
> ## Individual Comments
> 
> p1. "new research continuously emerge". To support this, it would be
> better to cite more recent papers in addition. The latest paper cited
> in this paragraph is published in 2015.

TODO

> p2. "<pal> ::= 0 | 1 | 0 <pal> 0 | 1 <pal> 1". Reading the following
> discussions, I suspect that the production of ϵ is missing and the
> correct one should be:
> 
>    <pal> ::= ϵ | 0 | 1 | 0 <pal> 0 | 1 <pal> 1

DONE

> p2. "The idea of automatic differentiation." I would feel that the
> idea here would be more similar to symbolic differentiation.

DONE

> p3. "... the recursive unfolding we performed ..." However, I don't
> think there is an issue in symbolic treatment, which can derive the
> grammar (d0 <rec>) ::= (d0 <rec>) from <rec> ::= <rec>. Both are valid
> CFGs.

I have clarified that we explain a naive derivation producedure based on recursive unfolding in this section.

> p3. "Another challenge ... is how to encode their recursive nature." I
> suspect this is still a challenge, because we already have a number of
> existing approaches of mechanized formalization of CFGs. Also, we can
> use a general technique that works for general DSLs, such as de Bruijn
> index and parametric HOAS.
> 
> This is in Haskell, but the following paper discusses CFGs represented
> by using de Bruijn index:
> 
>    Arthur Baars, S. Doaitse Swierstra, Marcos Viera: Typed
>    Transformations of Typed Grammars: The Left Corner Transform.
>    ENTCS 253 (7), 2010.  https://eur03.safelinks.protection.outlook.com/?url=https%3A%2F%2Fdoi.org%2F10.1016%2Fj.entcs.2010.08.031&data=05%7C02%7Cj.s.reinders%40tudelft.nl%7C4020b24233f24d08a73f08dd6c64e624%7C096e524d692940308cd38ab42de0887b%7C0%7C0%7C638785902816949725%7CUnknown%7CTWFpbGZsb3d8eyJFbXB0eU1hcGkiOnRydWUsIlYiOiIwLjAuMDAwMCIsIlAiOiJXaW4zMiIsIkFOIjoiTWFpbCIsIldUIjoyfQ%3D%3D%7C0%7C%7C%7C&sdata=knSvdZ8cwmqrVnAbyGZJFQk4MThzcmaUZ6bV5Q3xWbw%3D&reserved=0

I have simply removed this paragraph, because I don't think it is relevant.

@@

> p6. Explain "Dec.map" and "mk⇔".

TODO

> p10. "... we have learned the following three lessons." I suspect that
> at least the first two of the three lessons can be obtained
> straightforwardly from the existing results.

Done: I've rewritten this to make it clearer that these are not necessarily contributions of this paper.

> p15 "His work does cover mutually recursive nonterminals ..." He is
> discussing μ-regular expressions, which is not mutually recursive,
> whereas their languages are the same as the ones of CFGs.

TODO: μ-regular expressions can be mutually recursive. For example, the expr-stmt example in our introduction can be written like this:

expr = μ (\expr -> expr '+' expr | 0 | 1 | μ (\stmt -> expr | stmt * ';' * stmt))

We should perhaps make this clear in the discussion section

> p15. "... parsing is only the boolean ..." I think that this applies
> also to the proposed approach.

TODO: Our approach constructs a full parse tree. We do indeed produce only (at
most) a single parse tree and not all possible parse trees, but that does not
mean we only produce a boolean.

How can we explain this more clearly?

> p15. What are "μ-regular languages"? Anyway, I suspect that at least
> [13] discuss CFGs.

TODO: Indeed, we should explain clearly somewhere what μ-regular expressions are
and how they relate to context-free languages/grammars.

> p16. "... we have formalized (acyclic) context-free grammars ...". In
> my understanding, the proposed approach can handle cyclic ones like
> <rec> ::= <rec>.

Done: we have replace 'acyclic' with 'non mutually recursive'

> ----------------
>  REPORT
> ----------------
> 
> As it says on the tin, this paper addresses treating context free languages within a type-theoretic framework. Authors extend on existing work in formal languages mechanized in Agda and the line of work lifting derivatives to context-free languages.
> 
> Key technical contributions include defining a suitable fixed-point construction in Agda to safely encode recursive grammars, formalizing language derivatives to parse strings incrementally, and proving correctness of the parsing. Authors identify some present limitations of their approach and indicate those as directions for future work.
> 
> The background section is nice, gentle, readable. The whole thing is well written and explained. This is straight-down-the-middle nice line of work. "Here's some interesting pieces of work by some folk, but they haven't been put together in this way, and there's obviously a missing hole in our knowledge in this area, let me fix it." Overall, this work effectively fills a gap in formal parsing literature.  providing a clean, foundational formalization of context-free language parsing, type theoretically.
> 
> I have only minor notes.
> 
> Forgive me if this is silly, but we should explicitly include the empty string in the grammar; unless we do so, "0110" is not in the language of the given grammar; as written all productions are of odd-length strings.
> 
> The footnote from 2.1 appears on the next page, and that both got me confused looking for where it came from, and also confused as the return type of Parser, making me think "is this 'Set' deliberate here or not?" (and I think it's a mistake)

There's little we can do about the footnote without messing up other aspects of the typesetting. 
I have fixed the Set in the signature of Parser, that was indeed a mistake.

> 
> * small
> 
> ** "Doing so for the <pal> grammar" -> "Performing this unfolding for the <pal> grammar"
> 
> ** I know it's not what you mean, but what you said is that recursively unfolding the <pal> nonterminal produces the derivative of <pal0> wrt 1.
> 
> "Now, how do we take the derivative of the grammar <pal0> w.r.t. the next bit (1) in our string? [Say explicitly why it can't be done for <pal0> as written.] A simple solution [to the just-mentioned problem] is to recursively unfold the ⟨pal⟩ non-terminal. [After this unfolding, we can take the derivative]."
> 
> ** "see similarity between this type and the"
> 
> Repeat it. "similarity between the type (w : String) → Dec (P w) and". Remove the room for ambiguity.
> 
> ** speficying [specifying] and parsing context-free languages.

Fixed

@@

> ** from which our disired [desired]

Fixed

> ** an anonymous reviewer of a draft of this paper have [has]

Fixed

> ** the interested reader to Hocroft et al. [Hopcroft]
> 
> Use cleveref and you'll never again misspell an author in text.

Fixed
